Making Sequential Programs Deterministic

### 2019-03-28 design.tex
Edits in section 2.3

Our type system supports three types of polymorphism:  type
polymorphism, basetype polymorphism, and qualifier polymorphism.
\todo{Give an example for each.}
\begin{itemize}
\item
In \theDeterminismCheckerimplementation,
Type polymorphism is handled by Java's generics mechanism, which
\theDeterminismChecker fully supports, including class and method generics,
inference, etc.
\item
Basetype polymorphism is enabled by writing a type qualifier on a use of a
type variable.
\item
Java does not provide a syntax that can be used for qualifier polymorphism,
so we use a special type qualifier name, \<@PolyDet>.
A qualifier-polymorphic method \<m> with signature $\forall \kappa . \kappa \<int> \times \<Det boolean> \rightarrow
\kappa \<String>$ would be written in our dialect of Java as
\begin{Verbatim}
@PolyDet String m(@PolyDet int, @Det boolean)
\end{Verbatim}
Each use of \<@PolyDet> stands for a use of the qualifier variable
$\kappa$, and there is no need to declare the qualifier variable $\kappa$.
\TheDeterminismChecker currently supports
qualifier
polymorphism on methods but not on classes.
Therefore, \<@PolyDet> may be written in methods (signatures and bodies)
but not on fields.
\end{itemize}

The underlying checker-framework on which we build \theDeterminismChecker has support for
polymorphic qualifiers. When a user writes a polymorphic qualifier on a method signature or a type parameter,
it indicates that it could resolve to any type in the type system depending on how it is used.
In \theDeterminismChecker, we define a polymorphic qualifier \<PolyDet>.
One of the most common locations for a polymorphic qualifier is at a method signature.
Consider the following method declaration annotated with polymorphic qualifiers:
\begin{verbatim}
PolyDet int foo(PolyDet String param1, PolyDet boolean param2) { }
\end{verbatim}
\todo{It has the standard meaning.  There are two ways to interpret this.}
There are two valid instantiations for this declaration:
\begin{enumerate}
    \item @NonDet int foo(@NonDet String param1, @NonDet boolean param2) { }
    \item @Det int foo(@Det String param1, @Det boolean param2) { }
\end{enumerate}
This indicates that method \<foo> can be called with arguments having any of the \codeid{Det or NonDet} type qualifiers.
%   (\<@OrderNonDet> is not allowed because it is invalid on primitive types).
 \<PolyDet> resolves to the least upper bound of
the actual types on arguments. For instance, if this method is called as \codeid{foo(@Det String arg1, @NonDet boolean arg2)}, the 
checker framework resolves the method declaration as \codeid{@NonDet int foo(@NonDet String param1, @NonDet boolean param2)}
causing the return type to have the type qualifier \<NonDet>.

@@

Recall that \theDeterminismChecker does not permit \<OrderNonDet> on a type
that isn't a collection.
As a consequence, resolution of polymorphic type qualifiers at method signatures could have undesirable effects.
Suppose a method declaration is annotated as \codeid{PolyDet int size(PolyDet List<T> list)}. Invoking this method
with an argument of type \codeid{OrderNonDet List<Det String>} will result in \<PolyDet> getting resolved to \<OrderNonDet> which is invalid
on the return type \<int>. We circumvent this scenario by introducing two variants to the \<PolyDet> qualifier, \<PolyDet$\uparrow$>
and \<PolyDet$\downarrow$>.

@@

The semantics of these variants are as follows:
\begin{itemize}
    \item \<PolyDet$\uparrow$>: Same as \<PolyDet> except when it resolves to \<OrderNonDet>, it is replaced with \<NonDet>.
    \item \<PolyDet$\downarrow$>: Same as \<PolyDet> except when it resolves to \<OrderNonDet>, it is replaced with \<Det>.
\end{itemize} 
The modifiers $\uparrow$ and $\downarrow$ are indicative of the subtyping relationship of \<OrderNonDet> with \<NonDet> and \<Det> respectively
in the determinism type hierarchy.

@@

\begin{figure}
    $\infer[\textsc{poly resolution}]{\pi_r \ \beta_r\ \<m>(\overline{\pi_i \ \beta_i})\{\ \}}{\forall \kappa \in \{\|NonDet, OrderNonDet, Det|\} : \kappa \ \beta_r\ \<m>(\overline{K \ \beta_i})\{\ \} & \pi_r, \pi_i \in \{\|PolyDet|, \|PolyDet|\uparrow, \|PolyDet|\downarrow    \}}$
    \caption{Polymorphic resolution rule}
    \label{fig-poly-resolution}
\end{figure}

\begin{table}[]
    \begin{tabular}{|l|l|l|}
        \hline
        \textbf{\<PolyDet>} & \textbf{\<PolyDet$\uparrow$>} & \textbf{\<PolyDet$\downarrow$>} \\ \hline
        \<NonDet> & \<NonDet> &  \<NonDet>\\ \hline
        \<OrderNonDet> & \<NonDet> &  \<Det>\\ \hline
        \<Det> & \<Det> &  \<Det>\\ \hline
    \end{tabular}
\caption{Polymorphic resolution}
\todo{Write this as a rule rather than a table.}
\label{tab-poly-resolutions}
\end{table}

@@

\subsection{Discussion on defaulting rules}\label{defaulting}
Every type system in the Checker Framework must define a default type qualifier. This qualifier will be applied
to every location that isn't explicitly annotated. In \theDeterminismChecker, we chose \<PolyDet> as the default for all 
method parameters and return types including constructors.
The rationale for this design choice was to not make our type checker too restrictive by making everything \<Det>.  This implies that
\theDeterminismChecker still allows some non-determinism as long as it doesn't break any determinism specifications explicitly written by the programmer.

@@

\subsection{Dataflow analysis and type refinement}\label{dataflow}
The checker framework performs automatic type inference for local types within method bodies.
It also performs dataflow analysis and refines the types of these locals, thereby reducing the qualifier burden 
on the programmer. Consider the example below:
\begin{verbatim}
void foo(@Det int x) {
    @NonDet int y = x;    // Type of y gets refined to @Det.
    @Det int z = y;       //  No error here.
}
\end{verbatim}
In the example above, even though the local variable \codeid{y} was declared to be \codeid{NonDet}, it gets
type refined to \codeid{Det} after the first assignment statement. As the result, the assignment of \codeid{y}
to a \codeid{Det} variable \codeid{z} is sound. In the following example,
\begin{verbatim}
void foo(@NonDet Object x, @Det Object y) {
    x = y;                // No type refinement for x.
    @Det Object z = x;    // Error.
}
\end{verbatim}
\theDeterminismChecker reports a warning at the second assignment statement of \codeid{foo}. Type refinement only applies to locals. Refining non-locals can result in side effects making the analysis unsound.

With the above rules, the checker framework will automatically annotate a local collection with type \<NonDet> and its
type arguments will get the type qualifier \<Det>. For example, a locally declared \codeid{List} of \codeid{Strings} will be automatically annotated
as \codeid{NonDet List<Det String>}. Recall from~\ref{collection-rules} that \theDeterminismChecker does not allow this type on
any collection. While this is still sound, it will result in \theDeterminismChecker reporting a lot of errors because every locally declared 
collection gets annotated with this invalid type. To avoid this situation, we make an exception to the defaulting rules for local collections
and annotate their type parameters also with \<NonDet>. So a local list of Strings gets the default qualifier of \codeid{NonDet List<NonDet String>}. While this defaulting eliminates the possibility of automatically annotating locals collections with an invalid
type, it could result in a high number of false positives. In the following code snippet,
\begin{verbatim}
void test(@Det List<@Det String> argList) {
    List<String> localList = argList;
}    
\end{verbatim}
\theDeterminismChecker would report an error at the assignment statement since the inferred type for \codeid{localList} is 
\codeid{NonDet List<NonDet> String} and since collections types are invariant.
We therefore recommend that programmers explicitly annotate
all local variables that are collections to reduce false positives.

>>>>

Our type system supports three types of polymorphism:  type
polymorphism, basetype polymorphism, and qualifier polymorphism.
\todo{Give an example for each.}
\begin{itemize}
\item
In \theDeterminismCheckerImplementation,
type polymorphism is handled by Java's generics mechanism, which
\theDeterminismChecker fully supports, including class and method generics,
inference, etc.
Given the Java declaration
\begin{Verbatim}
<T> T identity(T arg) { return arg; }
\end{Verbatim}
the type of \<identity(arg)> is the same as the type of
\<arg>, and a programmer can write a type such as \codeid{List<@NonDet
  Integer>} (but see below for details about collections).
\item
Basetype polymorphism is enabled by writing a type qualifier on a use of a
type variable, which overrides the type qualifier at the instantiation
site.
For example, a \<choose> operation on sets could be defined in Java as
\begin{Verbatim}
class ChooseableSet<T> implements Set<T> {
  // return an arbitrary element from the set
  @NonDet T choose();
}
\end{Verbatim}
\item
Java does not provide a syntax that can be used for qualifier polymorphism,
so we use a special type qualifier name, \<@PolyDet>.
A qualifier-polymorphic method \<m> with signature $\forall \kappa . \kappa\ \<int> \times \<Det boolean> \rightarrow
\kappa\ \<String>$ would be written in our dialect of Java as
\begin{Verbatim}
@PolyDet String m(@PolyDet int, @Det boolean)
\end{Verbatim}
Each use of \<@PolyDet> stands for a use of the qualifier variable
$\kappa$, and there is no need to declare the qualifier variable $\kappa$.
\TheDeterminismChecker currently supports
qualifier
polymorphism on methods but not on classes.
Therefore, \<@PolyDet> may be written in methods (signatures and bodies)
but not on fields.
\end{itemize}

When a user writes a polymorphic qualifier on a method signature or a type parameter,
it indicates that it could resolve\todo{Should all occurrences of ``resolve'' be ``can be instantiated''} to any type in the type system depending on how it is used.
In \theDeterminismChecker, we define a polymorphic qualifier \<PolyDet>.
One of the most common locations for a polymorphic qualifier is at a method signature.
Consider the following declaration of addition:
\begin{verbatim}
@PolyDet int plus(@PolyDet int a, @PolyDet int b) { return a+b; }
\end{verbatim}
\todo{It has the standard meaning.  There are two ways to interpret this.}
There are two valid instantiations for this declaration:
\begin{enumerate}
    \item @NonDet int plus(@NonDet int a, @NonDet int b) { return a+b; }
    \item @Det int plus(@Det int a, @Det int b) { return a+b; }
\end{enumerate}
This indicates that method \<foo> can be called with arguments having any of the \codeid{Det or NonDet} type qualifiers.
%   (\<@OrderNonDet> is not allowed because it is invalid on primitive types).
 \<PolyDet> resolves to the least upper bound of
the actual types on arguments. For instance, if this method is called as \codeid{foo(@Det String arg1, @NonDet boolean arg2)}, \theDeterminismChecker resolves the method declaration as \codeid{@NonDet int foo(@NonDet String param1, @NonDet boolean param2)}
causing the return type to have the type qualifier \<NonDet>.

@@

Recall that \theDeterminismChecker does not permit \<OrderNonDet> on a type
that isn't a collection.
As a consequence, resolution of polymorphic type qualifiers at method signatures could have undesirable effects.
Suppose a method declaration is annotated as \codeid{PolyDet int size(PolyDet List<T> list)}. Invoking this method
with an argument of type \codeid{OrderNonDet List<Det String>} will result in \<PolyDet> getting resolved to \<OrderNonDet> which is invalid
on the return type \<int>. We circumvent this scenario by introducing two variants to the \<PolyDet> qualifier, \PolyDetUp
and \PolyDetDown.

@@

The semantics of these variants are as follows:
\begin{itemize}
    \item \PolyDetUp: Same as \<PolyDet> except when it resolves to \<OrderNonDet>, it is replaced with \<NonDet>.
    \item \PolyDetDown: Same as \<PolyDet> except when it resolves to \<OrderNonDet>, it is replaced with \<Det>.
\end{itemize} 
The modifiers $\uparrow$ and $\downarrow$ are indicative of the subtyping relationship of \<OrderNonDet> with \<NonDet> and \<Det> respectively
in the determinism type hierarchy.

@@

\begin{figure}
    $\infer[\textsc{poly resolution}]{\pi_r \ \beta_r\ \<m>(\overline{\pi_i \ \beta_i})\{\ \}}{\forall \kappa \in \{\|NonDet, OrderNonDet, Det|\} : \kappa \ \beta_r\ \<m>(\overline{K \ \beta_i})\{\ \} & \pi_r, \pi_i \in \{\|PolyDet|, \|PolyDet|\uparrow, \|PolyDet|\downarrow    \}}$
    \caption{Polymorphic resolution rule}
    \todo{I'm not sure what this is doing.  What code does it reject?}
    \todo{This suggests that \PolyDetUp is a qualifier, rather than
      $\uparrow$ being an operator on qualifiers.  We should clarify which
      it is.}
    \label{fig-poly-resolution}
\end{figure}

\begin{table}[]
    \begin{tabular}{|l|l|l|}
        \hline
        \textbf{\<PolyDet>} & \textbf{\PolyDetUp} & \textbf{\PolyDetDown} \\ \hline
        \<NonDet> & \<NonDet> &  \<NonDet>\\ \hline
        \<OrderNonDet> & \<NonDet> &  \<Det>\\ \hline
        \<Det> & \<Det> &  \<Det>\\ \hline
    \end{tabular}
\caption{Polymorphic resolution}
\todo{Write this as a rule rather than a table.}
\label{tab-poly-resolutions}
\end{table}

@@

\subsection{Discussion on defaulting rules}\label{defaulting}

\TheDeterminismChecker applies a default qualifier at each unannotated type
use (except uses of type variables, which already stand for a type that was
defaulted at the instantiation site).
This does not change the expressivity of the type system; it merely makes
the system more convenient to use.

In \theDeterminismChecker, we chose \<PolyDet> as the default for all 
method parameters and return types including constructors.
The rationale for this design choice was to not make our type checker too restrictive by making everything \<Det>.  This implies that
\theDeterminismChecker still allows some non-determinism as long as it doesn't break any determinism specifications explicitly written by the programmer.

@@

\subsection{Dataflow analysis and type refinement}\label{dataflow}
\TheDeterminismChecker performs automatic type inference for local types within method bodies.
It also performs dataflow analysis and refines the types of these locals, thereby reducing the qualifier burden 
on the programmer. Consider the example below:
\begin{verbatim}
void foo(@Det int x) {
    @NonDet int y = x;    // Type of y gets refined to @Det.
    @Det int z = y;       //  No error here.
}
\end{verbatim}
In the example above, even though the local variable \codeid{y} was declared to be \codeid{NonDet}, it gets
type refined to \codeid{Det} after the first assignment statement. As the result, the assignment of \codeid{y}
to a \codeid{Det} variable \codeid{z} is sound. In the following example,
\begin{verbatim}
void foo(@NonDet Object x, @Det Object y) {
    x = y;                // No type refinement for x.
    @Det Object z = x;    // Error.
}
\end{verbatim}
\theDeterminismChecker reports a warning at the second assignment statement of \codeid{foo}. Type refinement only applies to locals. Refining non-locals can result in side effects making the analysis unsound.

With the above rules, \theDeterminismChecker will automatically annotate a local collection with type \<NonDet> and its
type arguments will get the type qualifier \<Det>. For example, a locally declared \codeid{List} of \codeid{Strings} will be automatically annotated
as \codeid{NonDet List<Det String>}. Recall from~\ref{collection-rules} that \theDeterminismChecker does not allow this type on
any collection. While this is still sound, it will result in \theDeterminismChecker reporting a lot of errors because every locally declared 
collection gets annotated with this invalid type. To avoid this situation, we make an exception to the defaulting rules for local collections
and annotate their type parameters also with \<NonDet>. So a local list of Strings gets the default qualifier of \codeid{NonDet List<NonDet String>}. While this defaulting eliminates the possibility of automatically annotating locals collections with an invalid
type, it could result in a high number of false positives. In the following code snippet,
\begin{verbatim}
void test(@Det List<@Det String> argList) {
    List<String> localList = argList;
}    
\end{verbatim}
\theDeterminismChecker would report an error at the assignment statement since the inferred type for \codeid{localList} is 
\codeid{NonDet List<NonDet> String} and since collections types are invariant.
We therefore recommend that programmers explicitly annotate
all local variables that are collections to reduce false positives.
